---
title: "GWAS"
format: html
editor: visual
---

## Exercise: Visualizing GWAS Results with `qqman`

### Objective

The goal of this exercise is to learn how to visualize Genome-Wide Association Study (GWAS) results using the R package `qqman`. You will first familiarize yourself with the package's tutorial. Then, you will use `tidyverse` tools to load and format a "headerless" data file (`chrsnp_pval`) and apply your `qqman` knowledge to create and interpret Manhattan and Q-Q plots.

-   The R packages `qqman` and `tidyverse` installed. If you don't have them, run:

```{r}
install.packages("qqman") install.packages("tidyverse") 
```

### Part 1: The Tutorial Warm-Up

Before analyzing our own data, let's learn how `qqman` works.

1.  **Load the package:**

```{r}
library(qqman) 
```

2.  **Open the Tutorial (Vignette):** Run this command to open the official tutorial:

```{r}
vignette("qqman") 
```

3.  **Review and Run:**

    -   Read through the vignette, paying close attention to the **`manhattan()`** and **`qq()`** functions.

    -   Run the example code snippets in your R console using the built-in `gwasResults` dataset.

    -   **Crucially, identify the four column names `manhattan()` requires.**

